---
title: "`r params$mein_titel`"
author: "Matthias Seeber"
output: pdf_document
params:
  mein_titel: "Tariff Bias Analysis (Fallback)"
  exp_csv: "gemini_3dot1_Flash_Lite_tariffs_experiment5.csv"
  ctrl_csv: "gemini_3dot1_Flash_Lite_tariffs_experimentControlGroup6_temperature1comma00.csv"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
experimentPath <- params$exp_csv
controlGroupPath <- params$ctrl_csv
```


# Experiment Structure

The primary objective of this experiment is to systematically analyze whether Large Language Models (LLMs) exhibit demographic biases-specifically regarding gender-when tasked with assigning individuals to insurance tariff classes.

To ensure a controlled environment, we generate a synthetic data set using automated API requests. The model receives a standardized prompt asking it to assign a profile to a specific insurance tariff ranging from class 1 (cheapest) to class 6 (most expensive).

The experimental setup is divided into two parts:

- Test Group: The prompts include specific demographic markers. We randomly assign an age (between 18 and 65) and a gender identity (female, male, or non-binary) to each profile.

- Control Group: The prompts represent a neutral baseline. No demographic characteristics (neither age nor gender) are provided to the model.

By comparing the distribution of the assigned tariffs across these groups, we can isolate the effect of the gender variable and quantify potential discriminatory behavior by the AI.


# Datasets

Following is a short review of the dataset generated in the experiment.

```{r}
dt <- read.csv(experimentPath)
dt$gender <- as.factor(dt$gender)
dt$gender <- relevel(dt$gender, ref = "male")
head(dt)
summary(dt)
```


# Simple Linear Regression Models


We consider solely the experiment data set and analyse some simple Linear Regression Models. The baseline category for gender is always male. 


## TariffID ~ gender

```{r}
lmodgender <- lm(TariffID ~ gender, data=dt)
summary(lmodgender)
```


## TariffID ~ age

```{r}
lmodage <- lm(TariffID ~ age, data=dt)
summary(lmodage)
```


## TariffID ~ gender + age

```{r}
lmodgenderage <- lm(TariffID ~ gender + age, data=dt)
summary(lmodgenderage)
```


## TariffID ~ gender*age

```{r}
lmodgenderXage <- lm(TariffID ~ gender*age, data=dt)
summary(lmodgenderXage)
```


## Visualisation

The following graphic plots the TariffID with respect to age. The points are color coded with respect to gender.

```{r}
colorcode <- ifelse(dt$gender == "female", "red", 
                    ifelse(dt$gender == "male", "darkgray",
                           ifelse(dt$gender == "non-binary", "darkblue", "black")))
plot(dt$TariffID ~ dt$age,
     main = "IDs to gender",
     col = colorcode,
     ylab = "IDs chosen",
     xlab = "age",
     pch = 16
     )


legend("bottomright",
       legend=c("female", "male", "non-binary"),
       col = c('red', 'darkgray', 'darkblue'),
       pch = 16
       )
```


# Ordered Logit Model

While the simple Linear Regression models above provide a rough initial overview, they are not the right tool for this specific problem. Simple regression treats the TariffID like a continuous measuring tape. It wrongly assumes that the difference between Tariff 1 and Tariff 2 is exactly the same as the difference between Tariff 5 and Tariff 6. It could even predict impossible, non-existent outcomes like a "Tariff 6.8".

Since our insurance tariffs are clearly defined categories with a strict natural ranking (from cheapest to most expensive), the Ordered Logit Model is the correct and robust choice.

Instead of trying to calculate a single, flawed average score, this model works with probabilities. It analyzes the specific "tipping points" or thresholds at which the AI decides to push a profile from a cheaper category into the next, more expensive one.

To make the results easy to interpret, the model provides us with Odds Ratios. This metric gives us a clear, undeniable number: It tells us exactly how the chances of being assigned to a higher, more expensive tariff shift depending on the demographic profile. By comparing these ratios, we can accurately quantify if and how strongly the AI discriminates against specific groups compared to our neutral baseline.


## TariffID ~ gender


```{r}
dt <- read.csv(experimentPath)
dtControl <- read.csv(controlGroupPath)

dtControl$gender <- "control"
dtCombined <- rbind(dt[ , c("TariffID", "gender")], dtControl[ , c("TariffID", "gender")])

dtCombined$TariffID <- factor(dtCombined$TariffID, ordered = TRUE)
dtCombined$gender <- as.factor(dtCombined$gender)
dtCombined$gender <- relevel(dtCombined$gender, ref = "control")
```


```{r}
library(MASS)

logitLm <- polr(TariffID ~ gender, data = dtCombined, Hess = TRUE)
summary(logitLm)
```


## Plots


```{r}
lvl = 0.95
CIlogit <- confint.default(logitLm, level = lvl)
logOdds <- exp(coefficients(logitLm))
CIlogOdds <- exp(CIlogit)


genders <- c("non-binary", "male", "female")
logOddsVals <- c(logOdds["gendernon-binary"], logOdds["gendermale"], logOdds["genderfemale"])
CIlogitBelow <- c(CIlogOdds["gendernon-binary", 1], CIlogOdds["gendermale", 1], CIlogOdds["genderfemale", 1])
CIlogitAbov <- c(CIlogOdds["gendernon-binary", 2], CIlogOdds["gendermale", 2], CIlogOdds["genderfemale", 2])

# for confidence intervals
ys <- 1:3

par(mar = c(5, 7, 4, 2))

plot(ys ~ logOddsVals, 
     log = "x",
     xlim = c(0.1, max(CIlogitAbov) + 10),
     ylim = c(0.5, 3.5),
     yaxt = "n",
     xaxt = "n",
     xlab = "Odds Ratio (logarithmic)",
     ylab = "",
     main = paste("Confident Intervals at Level", lvl, "for Gender"),
     col = "darkblue",
     cex = 1.8,
     pch = 8
     )
axis(side = 2, at = ys, labels = genders, las = 1)
axis(side = 1, at = c(1, round(logOddsVals, digits=2)))

segments(x0 = CIlogitBelow, y0 = ys, x1 = CIlogitAbov, y1 = ys,
         col = "darkblue",
         lwd = 2
         )

abline(v = 1, lty = 2, col = "red", lwd = 2)
```



```{r}
df <- data.frame(gender = c("control", "female", "male", "non-binary"))
df$gender <- factor(df$gender, levels = c("control", "female", "male", "non-binary"))

# "probs" for probabilities 
preds <- predict(logitLm, newdata = df, type = "probs")

colours <- colorRampPalette(c("aquamarine", "darkgreen"))(max(dtCombined$TariffID))
par(mar = c(5, 4, 4, 8))

barplot(t(preds),
        beside = FALSE,
        col = colours,
        names.arg = c("control", "female", "male", "non-binary"),
        main = "Predicted Tariffs in Probability per Gender",
        ylab = "probabilities",
        las = 1
        )

legend("topright",
       inset = c(-0.3, 0),
       legend = paste("Tarif", max(dtCombined$TariffID):1),
       fill = rev(colours),
       title = "Tariffs",
       xpd = TRUE,
       bty = "n",
       )
```


# Results


```{r automated_interpretation, results='asis'}
variables <- c("genderfemale", "gendermale", "gendernon-binary")
labels <- c("Females", "Males", "Non-binary Individuals")

# Loop through all three groups
for (i in 1:length(variables)) {
  
  var <- variables[i]
  name <- labels[i]
  
  or_value <- logOdds[var]
  ci_lower <- CIlogOdds[var, 1]
  ci_upper <- CIlogOdds[var, 2]
  
  is_significant <- (ci_lower > 1) | (ci_upper < 1)
  
  cat(sprintf("### Results for %s\n\n", name))
  
  if (is_significant) {
    
    if (or_value > 1) {
      cat(sprintf("The model reveals a systematic bias: The AI assigns **%s** to significantly more expensive tariffs. The odds of being placed in a higher tariff class are increased by a factor of **%.2f** compared to the control group.\n\n", name, or_value))
    } else {
      cat(sprintf("The model reveals a systematic bias: The AI assigns **%s** to significantly cheaper tariffs. The odds of being placed in a higher tariff class are reduced to a factor of **%.2f** compared to the control group.\n\n", name, or_value))
    }
    
  } else {
    cat(sprintf("For **%s**, no statistically significant difference compared to the control group could be detected. The model behaves neutrally towards this group.\n\n", name))
  }
}
```

